A Mechanised Library for Generic (Co)Programming

### 2024-02-26 paper/paper.tex
I've made a bunch of changes which I don't have any idea right now what hey are.

\section{Introduction}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or productivity.
For recursion, this is done by trying to (automatically) infer which argument in
the recursive call gets smaller with respects to the original input argument.
For productivity, the algorithm checks that the (co)recursive call appears
directly under a constructor to make sure that this function always produce at
least one element after each recursive step.

@@

This has been a long-standing challenge to come up with a sound algorithm which
rejects all programs that do not terminate (soundness) and accepts as many
well-defined programs as possible (completeness). Our aim is to provide an
additional tool to the every-day programmer to be able to extend the range of
recursive programs we can write.
\mvol{What is the ``every-day programmer'' here? Are they working in Coq/Agda?}

@@

First we prove that the hylo equation has a unique solution
\[
  x = a \comp F(x) \comp c
\]
for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
recursion scheme by reducing it to a hylo. The following recursion schemes can
all be proven by reducing them to an hylo: folds, mutual recursive functions,
recursion schemes from comonads and their respective coinductive dual.
\mpav{Moved this from Section 3}

The primary goals of this work are:
\begin{enumerate}
  \item that the
  mechanisation of hylomorphisms and their laws should be axiom-free; and
  \item that Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \mintinline{OCaml}{Obj.magic}.
\end{enumerate}
In OCaml, \mintinline{OCaml}{Obj.magic} allows for unsafe casts between types.
Although the use of \mintinline{OCaml}{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\mintinline{OCaml}{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml code.
\mvol{This is good!}

@@

In order to abstract recursion patters we need to be able to talk about functors
which have the suitable fixed-point properties. Instances of these functors are
the polynomial ones which are in one-to-one correspondence with the so called
\emph{containers}~\cite{AbbottAG05}.
In this section we introduce containers

\subsection{Functors and Containers}
\label{sec:containers}
In order to To construct fixpoints of functors, these need to build
\emph{strictly positive types}. A common approach to construct such strictly
positive endofunctors is to use \emph{container functors}~\cite{AbbottAG05}.
However, reasoning about container and container functor equality requires to
consider both functional extensionality and heterogeneous equality. This is
problematic since, as stated in Section~\ref{sec:intro}, our goals to remain
axiom-free.  Furthermore, standard representations of containers in Coq will
lead to OCaml code that contains \mintinline{OCaml}{Obj.magic}. We will explain
why, and our mechanisation of containers in the remainder of this section.

A container is defined by a type of \emph{shapes}, and a \emph{family of
position types} indexed by shapes.
\begin{minted}{Coq}
Context (Shape : Set) (Pos : Shape -> Set).
\end{minted}
An \emph{extension} of this container is a functor, whose action on objects is
given by:
\begin{minted}{Coq}
Record App (X : Set) := { shape : Shape; contents : Pos shape -> X }.
\end{minted}
In our mechanisation, we use \mintinline{Coq}{App} to represent container
extensions.  Generally, any functor that is isomorphic to a container extension
is called a \emph{container functor}. However, in our work we will call
\emph{container functor} to the functors defined by a specific container
extension.

@@

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \mintinline{Coq}{LFixR}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, a catamorphism takes an
\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
\begin{minted}{Coq}
Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkApp sx (fun e => f (kx e)))
  end.
\end{minted}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{minted}{Coq}
Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
\end{minted}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \mintinline{Coq}{f} and
\mintinline{Coq}{cata g}.
\begin{minted}{Coq}
Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{minted}
This definition, however, prevents us from composing catamorphisms and
anamorphisms. To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{minted}{Coq}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{minted}
Catamorphisms from this initial algebra are denoted by \mintinline{Coq}{ccata},
and they can be used to define the isomorphism between 
\mintinline{Coq}{{x : GFix F | FinF x}} and
\mintinline{Coq}{LFix F}. In particular: 
\begin{minted}{Coq}
Lemma cata_ccata `{equiv A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{equiv A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{minted}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. However, if
anamorphisms produce finite trees, then we can also encode in Coq the notion of
\emph{recursive anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.
Recursive coalgebras are coalgebras whose anamorphism produces finite trees.
The following predicate captures this fact. Note how all the predicate states
is that \mintinline{Coq}{h} can only be applied finitely many times:
\begin{minted}{Coq}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{minted}
We show how this predicate is equivalent to the anamorphism of such coalgebras
producing finite trees:
\begin{minted}{Coq}
Lemma ana_rec_term `{equiv A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{minted}
Recursive anamorphisms are implemented in Coq recursively on the proof that
they can only be applied finitely many times (\mintinline{Coq}{RecF}).  Note
that such proofs will be erased when extracting code to OCaml. This is standard
practice when building general recursive functions in Coq\footnote{\dcas{This
technique had a name in the community, but I cannot remember. Something
predicates}}. The core definitions are the following:
\begin{minted}{Coq}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

Definition rana_f__ `{equiv A} (c : Coalg F A)
  : forall x : A, RecF c x -> LFix F
  := fix f (x : A) (H : RecF c x) :=
       let hx := c x in
       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
\end{minted}
Here, \mintinline{Coq}{RecF_inv} provides the structurally smaller element for
the Coq recursive definition. We then couple coalgebras with a proof that they
are recursive, and add a coercion from recursive coalgebras to coalgebras.
\begin{minted}{Coq}
Definition RCoalg F A = {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
\end{minted}
Similarly to the rest of the definitions, we prove that recursive anamorphisms
are respectful, and use this fact to build them as higher order proper
morphisms, and prove its universal law, i.e.\ that recursive anamorphisms are
the unique solution to the equation 
\mintinline{Coq}{f =e l_in \o fmap f \o h},
where \mintinline{Coq}{h} is a recursive coalgebra.
\begin{minted}{Coq}
Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...

@@

\subsection{Recursive Hylomorphisms}

Given an $F$-algebra and $F$-coalgebra, a solution to the equation $f = a \circ
F\; f \circ c$ is known as a hylomorphism. Unfortunately, this definition
cannot be accepted by Coq, since there is no structurally smaller argument to
do recursion, and there is no corecursive definition that is guaranteed to be
guarded, due to the use of an arbitrary algebra $a$. Furthermore, solutions to
the hylomorphism equation are \emph{not} guaranteed to be unique, or to even
exist. To address these issues, we mechanise \textbf{recursive hylomorphisms}
which, as we will see, are guaranteed to have a unique solution.

Hylomorphisms \emph{are} the unique solution to $f = a \circ F\; f \circ c$, if
$c$ is a recursive coalgebra. We show this by first providing the definition
for hylomorphisms in Coq, and then proving that any function satisfying this
equation is a hylomorphism. We call these hylomorphisms with a recursive
coalgebra, \textbf{recursive hylomorphisms}.
\begin{minted}{Coq}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H 
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{minted}
The definition of recursive hylomorphisms is recursive on the proof that the
anamorphism can only be applied finitely many times on any input. We again use
\mintinline{Coq}{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
we prove that \mintinline{Coq}{hylo_def} is respectful, and use this proof to
built the corresponding higher-order proper morphism:
\begin{minted}{Coq}
Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
\end{minted}
Finally, we show that recursive hylomorphisms are the unique solution to
the hylomorphism equation.
\begin{minted}{Coq}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{minted}
This result can be used to prove hylomorphism laws in a way that is analogous
to pen-and-paper program calculation techniques. We show the rule for 
hylomorphism fusion as an example:
\begin{minted}{Coq}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) 
  (g2 : Alg F C) (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
  : f2 \o hylo g1 h1 =e hylo g2 h1.
Proof.
  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
  rewrite hylo_univ.
  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
  rewrite fmap_comp.
  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
  rewrite compA.
  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
  rewrite <- E2.
  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
  rewrite <- compA.
  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
  rewrite <- compA.
  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
  rewrite (compA g1).
  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
  rewrite <- hylo_unr.
  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
  reflexivity.
Qed.
\end{minted}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}, where two consecutive recursive computations, one that builds a
data structure, and another one that consumes it, can be fused together into a
single recursive definition. This, in turn, allows us to prove that a recursive
hylomorphism is the composition of a catamorphism and a recursive anamorphism.
\begin{minted}{Coq}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.

Corollary cata_ana_hylo `(F : Cont Sh P) `{equiv A} `{equiv B}
  (g : Alg F B) (h : RCoalg F A)
: cata g \o rana h =e hylo g h.
\end{minted}
This result will allow us to \emph{calculate} programs by providing small, easy
to verify implementations, and combining them into efficient recursive
functions by \emph{using Coq's proofs of hylomorphism laws as optimisation
passes}.

\section{Extraction}

\dcas{Talk About this}
\begin{minted}{OCaml}
  let rec msort x = match x with
| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
     let (l0, r) = splitL x [] [] in
     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
\end{minted}

@@

\subsection{Program Calculation}

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs from
their naive counterparts through systematic program transformation. This area
has been extensively explored over the years. Tesson et al.\ demonstrated the
efficacy of leveraging Coq to establish an approach for implementing a robust
system dedicated to verifying the correctness of program transformations for
functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and
Emoto went further and\ldots~\cite{MurataE19}

\mvol{Need more general citations for program calculation.}

\subsection{Termination Checking}

Termination and productivity are non-trivial properties of programs, and various
methods have been proposed for checking that these properties hold. A common
approach is guardedness checking~\cite{10.5555/646535.695850}, which is a
\emph{syntactic} check that definitions avoid the introduction of
non-normalizable terms. However, some desirable definitions will not be accepted
by such checks.

In particular, the problem of \emph{nonstructural recursion} is
well-studied~\cite{BOVE_KRAUSS_SOZEAU_2016}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.

\subsubsection{Sized Types}

Another approach to certifying termination of recursive functions is the use of
\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
track/verify various properties of recursive programs, including productivity
and termination~\cite{HughesPS96}. The core idea of sized types is that types
express bounds on the sizes of recursive data structures. With this approach,
algorithms are implemented in the standard functional way, following the
divide-and-conquer pattern of splitting, recurring, and merging, with the
addition of the data types being indexed by a static approximation of the
relative \emph{size}. An advantage of this approach is that it allows for code
to be written in a natural way. There are costs, however. Programs must be
written in a way that accounts for the handling of size indexes, and support for
the size types must be added to the language.

This approach has been used to express nonstructurally recursive
algorithms in Agda: Copello et al.\ used sized types in a straightforward
formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.

\subsubsection{Divide and Conquer Recursion}

Abreu et al.\ propose\ldots~\cite{AbreuDHJMS23}. Their approach has similarities to what is presented here.

They differ in requiring the assumption of the functional extensionality axiom.
Additionally, they are concerned with the erasure of termination proofs but have
little discussion of what the resulting extracted code \emph{looks like} (that
is, whether the extracted OCaml code resembles the natural formulation of the
recursive function).
\mvol{We should find out for sure if they use \texttt{obj.magic}.}

>>>>

\section{Introduction}
\label{sec:intro}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or productivity.
For recursion, this is done by trying to (automatically) infer which argument in
the recursive call gets smaller with respects to the original input argument.
For productivity, the algorithm checks that the (co)recursive call appears
directly under a constructor to make sure that this function always produce at
least one element after each recursive step.

@@

This has been a long-standing challenge to come up with a sound algorithm which
rejects all programs that do not terminate (soundness) and accepts as many
well-defined programs as possible (completeness). Our aim is to provide an
additional tool to the every-day programmer to be able to extend the range of
recursive programs we can write.

@@

First we prove that the hylo equation has a unique solution
\[
  x = a \comp F(x) \comp c
\]
for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
recursion scheme by reducing it to a hylo. The following recursion schemes can
all be proven by reducing them to an hylo: folds, mutual recursive functions,
recursion schemes from comonads and their respective coinductive dual.
\mpav{Moved this from Section 3}
goals are: (1) the
mechanisation of hylomorphisms and their laws should be axiom-free; and (2) Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
proven that, for higher-order programs, simple interoperations can lead to
incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
extracted code will have stronger safety guarantees, and will allow the safe
interoperation with other OCaml code.

@@

In order to abstract recursion patters we need to be able to talk about functors
which have the suitable fixed-point properties, i.e. those who have an initial
algebra. Instances of these functors are the polynomial ones.  A common approach
to construct such functors is to use \emph{container}~\cite{AbbottAG05}.  In
this section we introduce the basics of polynomial functors as containers
(Section~\ref{sec:containers}).  However, reasoning about containers equality
will require us to consider both functional extensionality and heterogeneous
equality. As stated in previous sections~\ref{sec:intro}, this is rather
problematic since a stric requirement for this work is that we should remain
axiom-free. For this purpose in Section~\ref{sec:types-morphisms} we introduce
a custom equivalence relation on types in order to avoid using the functional
extensionality axiom.

\subsection{Functors and Containers}
\label{sec:containers}
A container is defined by a type of \emph{shapes} and a \emph{family of position
types} indexed by shapes.
\begin{minted}{Coq}
Context (Shape : Set) (Pos : Shape -> Set).
\end{minted}
An \emph{extension} of this container is a functor defined as follows:
\begin{minted}{Coq}
Record App (X : Set) := { shape : Shape; contents : Pos shape -> X }.
\end{minted}
with the obvious action on morphisms. Generally, any functor that is isomorphic
to a container extension is called a \emph{container functor}. However, in our
work we will call \emph{container functor} to the functors defined by a specific
container extension.

@@

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \mintinline{Coq}{LFixR}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, catamorphisms take an
\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
\begin{minted}{Coq}
Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkCont sx (fun e => f (kx e)))
  end.
\end{minted}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{minted}{Coq}
Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
\end{minted}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \mintinline{Coq}{f} and
\mintinline{Coq}{cata g}.
\begin{minted}{Coq}
Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{minted}
This definition, however, prevents us from composing catamorphisms and
anamorphisms. To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{minted}{Coq}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{minted}
Using this predicate, we can define the notion of \emph{recursive coalgebras}
as those that build finite trees. Recursive coalgebras are implemented in Coq
recursively on the proof that they produce finite trees.  Such proofs will be
erased when extracting code to OCaml. This is standard practice when building
general recursive functions in Coq\footnote{\dcas{This technique had a name in
the community, but I cannot remember. Something predicates}}.
\begin{minted}{Coq}
Notation rana_f__ h :=
  (fix f x H :=
    let hx := h x in
    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).

Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...

@@

\subsection{Recursive Hylos}

Recursive hylomorphisms are built from an algebra and a recursive coalgebra:
\begin{minted}{Coq}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
: forall (x : A), RecF c x -> B
:= fix f x H :=
  match c x as h0
        return
        (forall e : Pos (shape h0), RecF c (cont h0 e)) ->
        B
  with
  | MkCont s_x c_x => fun H => a (MkCont s_x (fun e => f (c_x e) (H e)))
  end (RecF_inv H).
\end{minted}

\begin{minted}{Coq}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{minted}

Recursive hylo universal property allows us to prove common optimisations:

\begin{minted}{Coq}
    Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
    (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
    : f2 \o hylo g1 h1 =e hylo g2 h1.
  Proof.
    rewrite hylo_univ.
    rewrite fmap_comp.
    rewrite compA.
    rewrite <- E2.
    rewrite <- compA.
    rewrite <- compA.
    rewrite (compA g1).
    rewrite <- hylo_unr.
    reflexivity.
  Qed.

  Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
    (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
    : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.

  Corollary cata_ana_hylo `(F : Cont Sh P) `{equiv A} `{equiv B}
    (g : Alg F B) (h : RCoalg F A)
  : cata g \o rana h =e hylo g h.
\end{minted}

@@

\subsection{Sized Types}

Another approach to certifying termination of nonstructurally recursive
functions is the use of \emph{sized types}. Sized types were introduced in
\cite{HughesPS96} as a way to track/verify various properties of recursive
programs, including productivity and termination. The core idea of sized types
is that types express bounds on the sizes of recursive data structures. This
approach has been used to express nonstructurally recursive algorithms in Agda:
\cite{CopelloTB14} used sized types in a straightforward formulation of
mergesort. Such an approach has also been used in MiniAgda~\cite{abs-1202-3496}.
